Reproducibility and Open Science in Economics

### 2025-03-12 article.tex
Version that I will submit

@@

Figure~\ref{fig:access} lists the distribution of the four levels of restrictions plus  the no-restriction category, for the \aerpapers{} AER articles in \aeryear{}, interacted with whether the authors offered to share the files with the data editor.\footnote{Not all offers to share data are taken up, depending on available resources and time. Appendix Table~\ref{tab:access_provision} shows the same numbers.}  
%
Past statistics note that about 40\% of articles rely on data that is not freely available \parencite{herbert_reproduce_2024,einav_economics_2014}. 
%
In \aeryear{}, slightly more than 50\% of all articles had no data restrictions at all (included the data in the replicatin package), with another 19\% being `very easy to obtain'. The combined percentage of 70\% is higher than the earlier long-run average.  More importantly, out of the 39 data sources that are difficult to obtain, about a third could actually be shared with the data editor. It is also worthwhile highlighting that the `moderately difficult to obtain' tends to be data from research data center networks that have fairly robust mechanisms of access, so that in principle and in practice, many researchers do have access to the same data used in these articles, even when the data editor may not have the time to access these data, and when it is not a legal option for the authors to share the data with the data editor. Arguably, the only data sources that are of immediate concern are data where access is uniquely attributable to the authors of the paper, or where access is no longer possible.

@@

While many economists use data collected by others, some are primary data generators or collectors, for instance through laboratory or field experiments, as well as surveys. \textbf{Data collection} software in this context are the survey software tools (Qualtrics, LimeSurvey, SurveyCTO) used to conduct surveys (including as part of lab or field experiments), as well as customized experimental software, such as oTree \citep{chen_otreeopen-source_2016} and z-tree \citep{fischbacher_z-tree_2021}.

@@

The landscape is even harder to assess for \textbf{data collection software}. Open source packages otree and z-tree are often used for lab experiments. otree is cited by over by between 1,400 (OpenAlex, as of 2025-01-29) and 2,400 articles (Google Scholar, as of 2025-01-29), z-Tree is cited between 9,000 (OpenAlex) and 12,000 times (Google Scholar), but it is not clear how to benchmark that, given poor software citation practices in economics. Lab as well as field experiments will often use online survey platforms to augment experiments, or as primary data collection tool. Qualtrics, one of the big commercial platforms for surveys, is mentioned (but not cited) over 200,000 times (Google Scholar). Open source alternative LimeSurvey is mentioned about 31,000 times.

(INSERT HERE SOME INFO FROM AEA ARTICLES).

@@

The second channel is through the \textbf{software instructions}. Conditional on having the right high-level software and the appropriate resources (see next section), most economics papers have openly accessible code. The advent of data and code availability policies most certainly has helped, but also reflected attitudes already present in the researcher community.\footnote{For a review of the history of data and code availability policies in economics, see \citet{vlaeminck_dawning_2021}.} While not the first, the AEA's policy was first implemented in 2005 \citep{bernanke_editorial_2004,american_economic_association_data_2005}, but the oldest replication package curated by the AEA accompanied an 1999 article \citep{frankel_does_1999,frankel_replication_1999} preceding the policy by several years.\footnote{I believe the oldest replication package in economics may be \citet{koenker_asymptotic_1988-1}, see \url{https://aeadataeditor.github.io/posts/2023-02-02-oldest-replication-package-jae}.} While the focus of availability policies was on the data, the code often came along for the ride, albeit not always in its most complete form. The advent of newer data and code availability policies \citep[for the AEA, ][]{AEA-announcement-July-2019,AEA-announcement-July-2019}, culminating for now in multiple journals having policies compliant with \ac{DCAS} \citep{koren_data_2022}, such as the AEA \citep{american_economic_association_data_2024} and

@@

In the template README published by myself and several other data editors in economics \parencite{templateREADMEv1.1}, we emphasize that a README should provide enough instructions for a reasonable person to re-implement the analysis described in the replication package. Here, authors need to take into account that cutting edge methods, including technology, may require more information and instruction than for standard methods. Authors likely do not need to describe how to run an analysis in Stata, given its ubiquitous use in economics, and the ease with which instructions can be found more generally, but they may need to provide detailed step-by-step information if the technology used is rare or bespoke. For instance, the emerging use of \acp{LLM} and \ac{AI} methods is far from the economic mainstream as of the writing of this article. Recent articles are still identifying ways economists scan actually use these tools, both for personal productivity \parencite{korinek_generative_2023} and as part of the technical toolkit \parencite{athey_machine_2019,dell_deep_2024}. 
%Only one of the articles submitted to any of the AEA journals in 2024 used these techniques (to be named, not yet published).

The most recent modern toolkits are not the only resource constraints that might restrict broad access. While it might be argued that the use of proprietary software is restrictive, it is one of many resource constraints that may be binding for some researchers. Researchers in lower-ranked (and lower-funded) research institutions, including in \acp{LMIC}, may well not have the funds to purchase proprietary software, but access to computers may be equally constraining. The template README requests information on the type of computer that was used by the original researcher, to provide a benchmark to future re-users. Acquiring access to sufficient memory (\ac{RAM}), storage, and use over time of those resources can be expensive, even when renting such resources in cloud environments (which very few researchers appear to be doing). Traditionally, that access may be embedded within a single purchased computer, which may have (in 2024) around 32GB of RAM, 1-2 TB of storage, and have 4-12 compute cores available exclusively to the owner. More complex analyses may require access to compute clusters (using hundreds or thousands of compute cores), very large storage arrays (measured in the two- to three-digit TB range), and may require up to 1024 GB of RAM. Cutting edge analyses may require specialized chips, such as one or more \acp{GPU}, or even a cluster of \acp{GPU}. We have observed analyses that may run data cleaning or data acquisition processes for months at a time.

@@

Some newer research requires vastly different types of resources. Studies using raw satellite data may require more than 10TB of data storage \citep{khachiyan_using_2022,khachiyan_data_2022}, may need more than 20,000 compute hours on a cloud provider \citep{rudik_optimal_2020,rudik_data_2020}, or the use of one \citep{dell_deep_2024,dell_data_2025} or dozens \citep{khachiyan_data_2022} \acp{GPU}.\footnote{As of January 2025, the type of \ac{GPU} used by \citet{dell_data_2025}, costs between USD 4500 and USD 7700, or between 2 and 7 times as much as a standard laptop.} But  access to the code and data for these papers is open: The AEA-related replication packages for these articles are licensed under a standard \ac{CC-BY} license. Some of the data not included in the \citet{dell_data_2025} replication package is on Huggingface, also under a \ac{CC-BY} license \citep{silcock_newswire_2024}.

@@

Concern should be raised, however, for any data that is subject to a gatekeeping mechanism, however objective, impartial, and lightweight it may be at this time, as such mechanisms can and do disappear.  A particular striking example, as of the writing of this article, is access to data from the USAID-funded \ac{DHS} program. Data access of such data was classified as `moderately easy to obtain' in Table~\ref{tab:categories} until February 2025, as it took only a day or two to obtain access subject to a lightweight data use agreement, a process my team at the AEA regularly went through to obtain data used by other researchers. In February 2025, the second Trump administration shut down USAID and the DHS program, with very little notice. While the system remained (for a time) accessible to those with prior access permission, no further access requests were accepted. That turns it into a `very difficult to obtain' dataset.\footnote{As of March 2025, efforts are underway to preserve the DHS program data, for instance via IPUMS, and to resuscitate both the access to historical data, as well continue funding new surveys.} These events are a note of caution that any kind of redistribution restriction may very well negatively impact future availability to the research community.

>>>>

Journals play a key role in this space, and will be an important background to my discussion and experiences, possibly also my biases. Most top economics journals have a data (and possibly code) availability policy.\footnote{For a review of the history of data and code availability policies in economics, see \citet{vlaeminck_dawning_2021}.} The AEA's policy was first implemented in 2005 \citep{bernanke_editorial_2004,american_economic_association_data_2005}. While the focus of early availability policies was on the data, the code often came along for the ride, albeit not always in its most complete form. I am the \ac{AEA}'s inaugural data editor, appointed in 2018 \citep{10.1257/pandp.108.745}. The AEA implemented a new policy in 2019 \citep{AEA-announcement-July-2019,AEA-announcement-July-2019}. Many other economics journals appointed data editors around that time, and  multiple journals coordinated on a common policy core called  \ac{DCAS} \citep{koren_data_2022}, and revised their policies to align with \ac{DCAS}, see \citet{american_economic_association_data_2024} for the AEA.\footnote{These initiatives are not restricted to economics, of course. Political science \citep{noauthor_data_2014,jacoby_american_2015,basile_research_2023}, sociology \citep{sociological_science_manuscript_2018,weeden_crisis_2023}, and general initiatives like the Transparency and Openness Promotion (TOP) guidelines \citep{nosek_promoting_2015}.} A key part of these newer policies was increased pre-publication monitoring of the content of replication packages \citep{10.1257/pandp.108.745,ChristianInt.J.Digit.Curation2018}.

@@

Figure~\ref{fig:access} lists the distribution of the four levels of restrictions plus  the no-restriction category, for the \aerpapers{} AER articles in \aeryear{}, interacted with whether the authors offered to share the files with the data editor.\footnote{Not all offers to share data are taken up, depending on available resources and time. Appendix Table~\ref{tab:access_provision} shows the same numbers.}  
%
Past studies have found that about between 40\% and 60\% of articles rely on data that are not freely available \parencite{herbert_reproduce_2024,hamermesh_six_2013,chetty_time_2012}. 
% Einav cite Hamermesh, claim 70% empirical papers, of which a "substantial majority  use data that have been assembled"
In \aeryear{}, slightly more than 50\% of all articles had no data restrictions at all (included the data in the replicatin package), with another 19\% being `very easy to obtain'. The combined percentage of 70\% is higher than the earlier long-run average.  More importantly, out of the 39 data sources that are difficult to obtain, about a third could  be shared with the data editor. It is also worthwhile highlighting that the `moderately difficult to obtain' tends to be data from research data center networks that have fairly robust mechanisms of access, so that in principle and in practice, many researchers do have access to the same data used in these articles, even when the data editor may not have the time to access these data, and when it is not a legal option for the authors to share the data with the data editor. Arguably, the only data sources that are of immediate concern are data where access is uniquely attributable to the authors of the paper, or where access is no longer possible.

@@

While many economists use data collected by others, some are primary data generators or collectors, for instance through laboratory or field experiments, as well as surveys. \textbf{Data collection} software in this context are the survey software tools (Qualtrics, LimeSurvey, SurveyCTO, KoboToolbox) used to conduct surveys (including as part of lab or field experiments), as well as customized experimental software, such as oTree \citep{chen_otreeopen-source_2016} and z-tree \citep{fischbacher_z-tree_2021}.

@@

The landscape is even harder to assess for \textbf{data collection software}. Open source packages otree and z-tree are often used for lab experiments. otree is cited  by between 1,400 (OpenAlex, as of 2025-01-29) and 2,400 articles (Google Scholar, as of 2025-01-29), z-Tree is cited between 9,000 (OpenAlex) and 12,000 times (Google Scholar), but it is not clear how to benchmark that, given poor software citation practices in economics. Lab as well as field experiments will often use online survey platforms to augment experiments, or as primary data collection tool. Qualtrics, one of the big commercial platforms for surveys, is mentioned (but not cited) over 200,000 times (Google Scholar). Open source alternative LimeSurvey is mentioned about 31,000 times.

%(INSERT HERE SOME INFO FROM AEA ARTICLES).

@@

The second channel is through  \textbf{software instructions}. Conditional on having the right high-level software and the appropriate resources (see next section), most economics papers have openly accessible code. The advent of data and code availability policies most certainly has helped, but also reflected attitudes already present in the researcher community. As noted earlier, the AEA's earliest policy was announced in 2004 \citep{bernanke_editorial_2004}, but the oldest replication package curated by the AEA accompanied an 1999 article \citep{frankel_does_1999,frankel_replication_1999} preceding the policy by several years.\footnote{The oldest  replication package in economics may be \citet{koenker_asymptotic_1988-1} in the JAE replication archive, see \url{https://aeadataeditor.github.io/posts/2023-02-02-oldest-replication-package-jae}, which actually contained data processing code, but not the data analysis code. That was only published in a later replication package, \citet{koenker_reproducible_2009}.} The newer policies \citep{AEA-announcement-July-2019,koren_data_2022} required that more code be provided (starting with raw, rather than cleaned data), and the systematic provision of supplementary materials, such as the survey code to use with data collection software. These policies require that the code be made available in a reasonably liberal license, though no specific open source license is specified. Furthermore, most of the top journal's repositories of replication packages are at trusted repositories, and not behind journal paywalls. Thus, software instructions in economics, as associated with the scientific output in journals, is generally cost-free, and almost always available under an open access license.

@@

In the template README published by myself and several other data editors in economics \parencite{templateREADMEv1.1}, we emphasize that a README should provide enough instructions for a reasonable person to re-implement the analysis described in the replication package. Authors need to take into account that cutting edge methods, including technology, may require more information and instruction than for standard methods. Authors likely do not need to describe how to run an analysis in Stata, given its ubiquitous use in economics, and the ease with which instructions can be found more generally, but they may need to provide detailed step-by-step information if the technology used is rare or bespoke. For instance, the emerging use of \acp{LLM} and \ac{AI} methods is far from the economic mainstream as of the writing of this article. Recent articles are still identifying ways economists scan actually use these tools, both for personal productivity \parencite{korinek_generative_2023} and as part of the technical toolkit \parencite{athey_machine_2019,dell_deep_2024}. 
%Only one of the articles submitted to any of the AEA journals in 2024 used these techniques (to be named, not yet published).

The most recent modern toolkits are not the only resource constraints that might restrict broad access. While it might be argued that the use of proprietary software is restrictive, it is one of many resource constraints that can be binding for some researchers. Researchers in lower-ranked (and lower-funded) research institutions, including in \acp{LMIC}, may well not have the funds to purchase proprietary software, but access to computers may be equally constraining. The template README requests information on the type of computer that was used by the original researcher, to provide a benchmark to future re-users. Acquiring access to sufficient memory (\ac{RAM}), storage, and use over time of those resources can be expensive, even when renting such resources in cloud environments (which very few researchers appear to be doing). Traditionally, that access may be embedded within a single purchased computer, which may have (in 2024) around 32GB of RAM, 1-2 TB of storage, and have 4-12 compute cores available exclusively to the owner. More complex analyses may require access to shared compute clusters (using hundreds or thousands of compute cores), very large storage arrays (measured in the two- to three-digit TB range), and may require up to 1024 GB of RAM. Cutting edge analyses may require specialized chips, such as one or more \acp{GPU}, or even a cluster of \acp{GPU}. I have observed analyses that may run data cleaning or data acquisition processes for months at a time.

@@

Some newer research requires vastly different types of resources. Studies using raw satellite data may require more than 10TB of data storage \citep{khachiyan_using_2022,khachiyan_data_2022}, may need more than 20,000 compute hours on a cloud provider \citep{rudik_optimal_2020,rudik_data_2020}, or the use of one \citep{dell_deep_2024,dell_data_2025} or dozens \citep{khachiyan_data_2022} \acp{GPU}.\footnote{As of January 2025, the type of \ac{GPU} used by \citet{dell_data_2025}, costs between USD 4500 and USD 7700, or between 2 and 7 times as much as a standard laptop.} Access to the code and data for the papers mentioned is open: The AEA-related replication packages for these articles are licensed under a standard \ac{CC-BY} license. Some of the data not included in the \citet{dell_data_2025} replication package is on Huggingface, also under a \ac{CC-BY} license \citep{silcock_newswire_2024}. Open access to satellite data is one of the canonical examples of the benefits of open access \citep{nagaraj_improving_2020}. In these cases, the computational resources may restrict the benefits of the open access of data and code.

@@

I have described in this article how open data and code are in the academic literature in economics, and how access to software and hardware resources can be limiting factors. Almost all code is openly accessible in top journals. The vast majority of data is accessible with little to no effort, and a large proportion of the remaining restricted data can be accessed by networks that include thousands of researchers. I provide a few concrete examples where the openness of the data and code available allows others to directly build on prior results. Broader assessments of the benefits of open access are more difficult to measure, in part because the right controls are hard to construct, in part because the community is still only starting to learn how to technically leverage the openness in a large-scale fashion.

Nevertheless, access to networks of data access and financing remains one of the key worries. I am a regular participant in discussions within three research networks that provide access to restriced data (\ac{FSRDC}, \ac{CRDCN}, and \ac{CASD}). Core discussions center around equity and access, and how to balance those criteria while preserving the privacy of the respondents for which these networks act as curators.

One under-appreciated aspect of open access is that it enables persistence. 
For any data that is subject to a gatekeeping mechanism, however objective, impartial, and lightweight it may be at the present time, such mechanisms can and do disappear. Every one of the networks that I mention above has a mandate to work within budget constraints, and those budgets are determined fundamentally by external forces, typically government-based funding agencies. A particular striking example, as of the writing of this article, is access to data from the USAID-funded \ac{DHS} program. Data access to DHS data was classified as `moderately easy to obtain' (as per Table~\ref{tab:categories}) until February 2025, as it took only a day or two to obtain access subject to a lightweight data use agreement. My team at the AEA regularly went through the process to obtain data used by other researchers, a process that was easy to navigate even for an undergraduate researcher on my team. In February 2025, the second Trump administration shut down USAID and ``paused'' the DHS program, with very little notice, and no recourse. While the DHS program system was still accessible to those with prior access permission in late February 2025, no further access requests were accepted. That turns it into a `very difficult to obtain' dataset.\footnote{As of March 2025, efforts are underway to preserve the DHS program data, for instance via IPUMS, and to resuscitate both the access to historical data, as well continue funding new surveys.} These events are a note of caution that any kind of redistribution restriction may very well negatively impact future availability to the research community. Whereas journals can subscribe to mechanisms that allow past issues to remain accessible even when the journal is shuttered,\footnote{See \href{https://clockss.org/about/how-clockss-works/}{CLOCKSS}.} and open access software can be preserved by communities with an interest in continued use,\footnote{See f.i. \href{https://www.datarescueproject.org/}{Data Rescue Project}.} no such mechanism exists for most restricted-access data. While open code may ensure we can recompute, and advances in computational infrastructure will bring the current cutting edge into the space of everyday-accessibility, data that are not open can and will disappear. That is concerning.
